Follow-up Samples

### 2017-01-19 submission/manuscript_outline_20161024.Rmd
streamline -discussion

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***FIT and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

@@

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance.  Data columns that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyperparameter involved taking the samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created.

***Initial Follow Up Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort follow up data set was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.

***Selection of Important OTUs***
In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or initial follow up models.

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in the lesion model were then analyzed using a paired wilcoxson test.  We then investigated the relative abundance of specific previously CRC associated bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial follow up model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

***Reproducible methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.  Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005).

>>>>

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***FIT and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

@@

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyperparameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyperparameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Before Sample Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.

***Selection of Important OTUs***
In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or before sample models.

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in either model were then analyzed using a paired wilcoxson test.  We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the before sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature before sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

***Reproducible Methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.  Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

### 2017-01-19 submission/manuscript_outline_20161024.tex
streamline -discussion

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}21{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described previoulsy but also a follow up
sample between 188 - 546 days after surgery and treatment had been
completed. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}23{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with control, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

@@

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance. Data columns that had near zero variance were removed.
This pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyperparameter involved taking the samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
through 20 repeated 10-fold cross validations to optimzie the model's
AUC (Area Under the Curve of the Receiver Operator Characteristic). This
resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. Once the ideal mtry was
found the entire 490 sample set was used to create the final Random
Forest model on which testing on the 67-person cohort was completed. The
default cutoff of 0.5 was used as the threshold to classify individuals
as positive or negative for lesion. The hyperparameter, mtry, defines
the number of variables to investigate at each split before a new
division of the data is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The main difference was that only the 67-person cohort
follow up data set was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyperparameter was completed using the same
procedure that was used to create the lesion model.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most important to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion or initial follow up models.

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}25{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
wilcoxson ranked sum test. Where multiple comparison testing was needed
a Benjamini-Hochberg (BH) correction was applied {[}26{]} and a
corrected P-value of less than 0.05 was considered significant. Unless
otherwise stated the P-values reported are those of that were BH
corrected.

\textbf{\emph{Analysis Overview}} Initial and follow up samples were
analyzed for differences in alpha and beta diversity. Next, differences
in FIT between initial and follow ups for either adenoma or carcinoma
were investigated. From here, all OTUs that were used in the lesion
model were then analyzed using a paired wilcoxson test. We then
investigated the relative abundance of specific previously CRC
associated bacteria, specifically, OTUs that taxonomically classified to
\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}. We wanted to test if there were any differences based on
whether the individual had an adenoma or carcinoma. From here the lesion
model was then tested for accuracy in prediction and whether it reduced
the positive probability of lesion after surgery. The most important
OTUs for this were used to build a reduced model and it was assessed for
similarity to the original model. We then used the initial follow up
model to assess whether it could classify samples better then the lesion
model. The most important OTUs were then identified from this model and
used to create a reduced feature initial follow up model. This reduced
feature model, as was done with the lesion model, was compared to the
full model for loss of accuracy. Finally, a list of common OTUs were
found for the two different models used.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005).

>>>>

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to normalize within this time
period {[}21{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described but also a follow up sample
between 188 - 546 days after surgery and treatment had been completed.
This study was approved by the University of Michigan Institutional
Review Board. All study participants provided informed consent and the
study itself conformed to the guidelines set out by the Helsinki
Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}23{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with normal, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

@@

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance and OTUs that had near zero variance were removed. This
pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyperparameter involved taking the samples and
making 100 different 80/20 (train/test) splits of the data where normal
and lesion were represented in the same proportion within both the whole
data set and the 80/20 split. Each of these splits were then run through
20 repeated 10-fold cross validations to optimize the mtry
hyperparameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 490 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyperparameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

\textbf{\emph{Before Sample Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The main difference was that only the 67-person cohort
was used at all stages of model building and classification. Other than
this difference the creation of this model and optimization of the mtry
hyperparameter was completed using the same procedure that was used to
create the lesion model. Instead of classifying samples as positive or
negative of lesion this model classified samples as positive or negative
for being a before surgery sample.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most important to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different 80/20 split models and then
filtered this list to variables that were only present more than 50\% of
the time. This final collated list of variables was what was considered
the most important for the lesion or before sample models.

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}25{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
wilcoxson ranked sum test. Where multiple comparison testing was needed
a Benjamini-Hochberg (BH) correction was applied {[}26{]} and a
corrected P-value of less than 0.05 was considered significant. Unless
otherwise stated the P-values reported are those that were BH corrected.

\textbf{\emph{Analysis Overview}} Initial and follow up samples were
analyzed for differences in alpha and beta diversity. Next, differences
in FIT between initial and follow ups for either adenoma or carcinoma
were investigated. From here, all OTUs that were used in either model
were then analyzed using a paired wilcoxson test. We then investigated
the relative abundance of specific previously associated CRC bacteria,
specifically, OTUs that taxonomically classified to \emph{Fusobacterium
nucleatum}, \emph{Parvimonas micra}, \emph{Peptostreptococcus
assacharolytica}, and \emph{Porphyromonas stomatis}. We wanted to test
if there were any differences based on whether the individual had an
adenoma or carcinoma. From here the lesion model was then tested for
accuracy in prediction and whether it reduced the positive probability
of lesion after surgery. The most important OTUs for this were used to
build a reduced model and it was assessed for similarity to the original
model. We then used the before sample model to assess whether it could
classify samples better then the lesion model. The most important OTUs
were then identified from this model and used to create a reduced
feature before sample model. This reduced feature model, as was done
with the lesion model, was compared to the full model for loss of
accuracy. Finally, a list of common OTUs were found for the two
different models used.

\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005) and
the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.
